---
title: "Panel Data Basics"
format: html
execute: 
  warning: false
---


<style>
.quarto-title .title {
  text-align: center;
}
</style>


## Packages

```{r}

library(tidyverse)
library(plm)
library(broom)

set.seed(1)

```


## Model - Random Effects
Assume we have the following panel data model:

$$
Y_{it} = \alpha + X_{it} \beta + \eta_i + \varepsilon_{it}
$$

where:

- $Y_{it}$ is the dependent variable for observation $i$ at time $t$  
- $X_{it}$ is the independent variable  
- $\alpha$ is a constant  
- $\beta$ is the coefficient of interest  
- $\eta_i \sim \mathcal{N}(0, \sigma_\eta^2)$ is the individual-specific effect  
- $\varepsilon_{it} \sim \mathcal{N}(0, \sigma_\varepsilon^2)$ is the idiosyncratic error term

A key assumption here is that the individual-specific effect is uncorrelated with the covariates:
$$
\text{Cov}(X_{it}, \eta_i) = 0
$$


## Simulated Data - Random Effects

```{r}

# Parameter Values
N_values <- c(10, 25, 50, 100, 200, 500, 1000)
T <- 5

alpha <- 1
beta <- 10

sigma_eta <- 1
sigma_eps <- 1

```

```{r}

# function to simulate and estimate models
simulate_and_estimate <- function(N) {
  
  id <- rep(1:N, each = T)
  time <- rep(1:T, times = N)
  
  X <- rnorm(N * T)
  eta <- rep(rnorm(N, 0, sigma_eta), each = T)
  eps <- rnorm(N * T, 0, sigma_eps)
  Y <- alpha + beta * X + eta + eps
  
  data <- tibble(id = factor(id), time, X, Y)
  
  ols <- lm(Y ~ X, data = data)
  re  <- plm(Y ~ X, data = data, index = c("id", "time"), model = "random")
  
  bind_rows(
    tidy(ols, conf.int = TRUE) |> mutate(model = "OLS", N = N),
    tidy(re, conf.int = TRUE)  |> mutate(model = "Random Effects", N = N)
  )
}


```

```{r}

# run simulations
results <- map_dfr(N_values, simulate_and_estimate)

```

```{r}

results %>% 
  filter(term=="X") %>% 
  ggplot(aes(x = N, y = estimate, color = model)) +
  geom_point() +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.1
  ) +
  scale_x_continuous(trans = "log10", breaks = N_values) +
  geom_hline(yintercept = beta, linetype = "dashed", color = "black") +
  labs(
    title = "Coefficient Estimates with 95% Confidence Intervals",
    subtitle = "Dashed line indicates true value of β = 10",
    y = "Estimated Coefficient", x = "Number of Individuals (N)"
  ) +
  theme_minimal()

```

```{r}

results %>% 
  filter(term=="(Intercept)") %>% 
  ggplot(aes(x = N, y = estimate, color = model)) +
  geom_point() +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.1
  ) +
  scale_x_continuous(trans = "log10", breaks = N_values) +
  geom_hline(yintercept = alpha, linetype = "dashed", color = "black") +
  labs(
    title = "Coefficient Estimates with 95% Confidence Intervals",
    subtitle = "Dashed line indicates true value of α = 1",
    y = "Estimated Coefficient", x = "Number of Individuals (N)"
  ) +
  theme_minimal()

```

